Models, derivatives, response diversity

### 2024-07-05 models_derivatives_response_diversity.Rmd
separate rmd for models, derivatives and response diversity calculattion

---
title: "manuscript_analysis"
author: "Sophie Moreau"
date: "2024-07-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(here)
library(readr)
library(viridis)

#Introduction
This is the supplementary information of the Master thesis. This document
covers the whole analysis performed to assess response diversity between fish communities
in Swiss peri-alpine lakes. 
This document has two parts: Data preparation (1) and Data analysis (2). 
Function to model general additive models is provided separately (see function_GAM.R). 
Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R)

```{r}
# do we want to include zurich and constance separately?
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)
projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)

projet_lac_data |> 
  distinct(LakeBasin)
```

#1. Data preparation
##Abundance data
```{r}
##read in the data: Simstrat model data and Projet Lac combined
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)

# preparing additional column with abundance data of each species
abundance_data <- projet_lac_data |> 
  # exclude rows without taxa information
  filter(!Taxa_latin_FINAL %in% "NO_FISH") |> 
  # exclude samples obtained by electro-fishing
  filter(!Protocol %in% "electro") |> 
  #select columns relevant for species individual occurrences
 group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Weightg_raw, Weightg_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |> 
  # prepare column with abundance of each species
  tally() |> 
  # values_fill = 0 added to make all NAs into 0
  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0)

# abundance of each species
abundance_data_long <- abundance_data |> 
  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")

# add presence-absence column
abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)

# View(abundance_data_long)

max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
max(abundance_data_long$Abundance)

#we only need information on species present in the lake, not all of them
abundance_data_long_lakepresence <- abundance_data_long |> 
  group_by(Lake, Species) |>
  # . stands for data being piped in do function
  do(LakePresence = max(.$Presence)) |> 
  unnest(cols = LakePresence)

# double-check 
abundance_data_long_lakepresence |>
  filter(LakePresence == "1") |> 
  group_by(Lake, LakePresence) |>
  count()

# join the two dfs
df_final <- abundance_data_long_lakepresence |> 
  filter(LakePresence == "1") |> 
  left_join(abundance_data_long )

#double-check if all lakepresence = 1
max(df_final$LakePresence)
min(df_final$LakePresence)

# save final df
saveRDS(df_final, "df_final.rds")

##Taxonomy and requirements
We pooled several species into ecomorphs to prevent loss of data-points. 
We also excluded species that were caught less than 10 times in the whole Projet Lac.

```{r}
# read in prepared df
df_final <- readRDS("df_final.rds")

#######refining errors in species' identity that were noticed upon examination of the
# data frame, following the table 16 in the Projet Lac "Synthesis report"

# C. brienzii should not be present in lake Thun, add to Coregonus sp.

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii") & Lake %in% c("Thun"),
                          "Coregonus_sp", as.character(Species)))

# Salvelinus umbla in Lake Thun should be Salvelinus sp.

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_umbla") & Lake %in% c("Thun"),
                          "Salvelinus_sp", as.character(Species)))

# Squalius squalus in Biel and Neuchatel to Squalius cephalus

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Squalius_squalus") & Lake %in% c("Biel", 
                                                                           "Neuchatel"),
                          "Squalius_cephalus", as.character(Species)))

#######Pooling species to be able to include more samples
# Profundal Cottus gobio are pooled

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c(
    "Cottus_gobio_Profundal_Thun",
    "Cottus_gobio_Profundal_Lucerne", "Cottus_gobio_Profundal_Walen"
  ),
  "Cottus_sp_Profundal", as.character(Species)
  ))

# Salvelinus is pooled into "profundal" and "limnetic" habitats

# profundal

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c( "Salvelinus_sp_Profundal_dwarf_Thun", 
                                          "Salvelinus_sp_Profundal_dwarf_VWS", 
                                          "Salvelinus_sp_Profundal_extreme_Thun",
    "Salvelinus_sp_Profundal_Walen_I", "Salvelinus_sp_Profundal_Walen_II"),
  "Salvelinus_sp_Profundal", as.character(Species)
  ))

# limnetic 
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_sp_Limnetic_Thun", "Salvelinus_sp_Limnetic_VWS"),
                          "Salvelinus_sp_Limnetic", as.character(Species)))

# Coregonus species are pooled to new groups based on ecomorphs,
# following De-Dayne et al. 2022

# Albeli
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_albellus", "Coregonus_candidus", "Coregonus_confusus",
                                         "Coregonus_heglingus", "Coregonus_zugensis"),
                          "Coregonus_sp_albeli", as.character(Species)))
# Balchen
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_alpinus", "Coregonus_arenicolus", "Coregonus_duplex",
                        "Coregonus_helveticus", "Coregonus_palaea"),
                          "Coregonus_sp_balchen", as.character(Species)))

# Felchen

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii", "Coregonus_fatioi", "Coregonus_intermundia",
                                         "Coregonus_litoralis", "Coregonus_macrophthalmus", "Coregonus_zuerichensis"),
                          "Coregonus_sp_felchen", as.character(Species)))
# Large pelagic

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_acrinasus", "Coregonus_wartmanni"),
                          "Coregonus_sp_large_pelagic", as.character(Species)))

# Benthic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_profundus",
                          "Coregonus_sp_benthic_profundal", as.character(Species)))
# Pelagic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_nobilis",
                          "Coregonus_sp_pelagic_profundal", as.character(Species)))

# "Coregonus_sarnensis": could not be assigned to Albeli or Felchen

# check species list
species <- df_final |> 
  distinct(Species)

# check how many species have less than 10 overall observations: 30
df_final |> 
  group_by(Species) |> 
  summarize(tot_abu = sum(Presence)) |> 
  filter(tot_abu <= 10)

# exclusion of species with less than 10 overall observations
df_models <- df_final |> 
  group_by(Species) |> 
  mutate(tot_abu = sum(Presence)) |> 
  ungroup() |> 
  filter(tot_abu >= 10)

# save the df of the refined species, df will be used for the models
saveRDS(df_models, "df_models.rds")

df_models$Species <- as.factor(df_models$Species)
levels(df_models$Species)

species <- df_models |> 
  group_by(Species) |> 
  summarize(tot_obs = sum(Abundance))

```
##Data frames for GAMs
We compiled four data frames for the four types of general additive models that 
were run: Species with 1) binomial dataoccurring in one lake,
2) abundance data occurring in one lake, 3) binomial data 
occuring in multiple lakes, 4) abundance data and occuring in multiple lakes.

```{r}
###Overview
#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
#2. species that occur only in one lake: Lake not added as random effect

#1. binomial vs. abundance data

df_models$Species <- as.character(df_models$Species)

non_binomial_species <- df_models |> 
  filter(Abundance > 1) |> 
  group_by(Species) |>
  distinct(Species) |> 
  pull(Species)

binomial_species <- df_models |> 
  filter(!Species %in% non_binomial_species) |> 
  distinct(Species) |> 
  pull(Species)

#2. binomial vs. abundance species occuring only in one lake

#1: GAM without lake as random effect, binomial

bi_one_occurence <- df_models |> 
  filter(Species %in% binomial_species) |>
  group_by(Species) |>
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)

# save df
df_bi <- df_models |>
  filter(Species %in% bi_one_occurence)

saveRDS(df_bi, "data_frame_models/df_binomial_gam")

#2 GAM abundance without lake as random effect
abu_one_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)

df_abu <- df_models |> 
  filter(Species %in% abu_one_occurence)

saveRDS(df_abu, "data_frame_models/df_abundance_gam")

#binomial vs. abundance species occuring in several lakes

#3 gam with random intercept, binomial
bi_multi_occurence <- df_models |> 
  filter(Species %in% binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_binomial_re <- df_models |> 
  filter(Species %in% bi_multi_occurence)

saveRDS(df_binomial_re, "data_frame_models/df_binomial_re")

#4 abundance GAM with random intercept
abu_multi_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_abundance_re <- df_models |> 
  filter(Species %in% abu_multi_occurence)

saveRDS(df_abundance_re, "data_frame_models/df_abundance_re")

#2. Data Analysis
##General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@

# function for the GAMs (old)
# source(here("function_GAM.R"))

# new function for temp and depth models 
source(here("function_temp_depth.R"))
```

@@

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions

# DO NOT RUN NOW, RUN LATER!
predictions_depth_temp(df_1)
predictions_depth_temp(df_2)
predictions_depth_temp(df_3)
predictions_depth_temp(df_4)

@@

@@

  model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                   s(Depth_sample, k = 3)
                            + s(fLake, bs = 're')
                            +  s(fProtocol, bs = 're'), family = ziP())
  print(glance(model_0))

@@

# 8 models are run with binomial temp and depth

@@

##Drivers of thermal response diversity

Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
```{r}
# Df with properties of lakes
lake_info <- read_xlsx("lake_info.xlsx")

# select the columns we use
lake_selection <- lake_info |> 
  select(Lake, Phos_max, Max_depth, Lake_area)

# information about categorization in each lake for the subgroups
species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
  rename(endemism = detail_category)

# exclude lepomis gibbosus and re-calculate species richness per lake
species_endemism_short <- species_endemism |> 
  filter(species != "Lepomis_gibbosus") |> 
  select(-sum_species) |> 
  group_by(fLake) |> 
  mutate(sum_species = sum(num_species))

species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)

str(species_endemism_short)

# adding column with number of species per category in each lake
df_species_endemism <- species_endemism_short |> 
  select(-species, - num_species) |> 
  add_count(fLake, endemism) |> 
  rename(count = n) |> 
  distinct(fLake, count, sum_species, endemism)

str(df_species_endemism)

df_species_endemism$count <- as.numeric(df_species_endemism$count)

# making it wide 
df_species_endemism_long <- df_species_endemism |> 
  pivot_wider(names_from = endemism, values_from = count) |> 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
  rename(Lake = fLake)

str(df_species_endemism_long)

# merge endemism info and lake info

endemism_lake <- merge(df_species_endemism_long, lake_selection)

# Read in data on response diversity metrics in each lake
resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
  select(temp:fLake)

# select response diversity columns we need
rdiv <- resp_div_all |>  
  select(temp, fLake, rdiv, sign) |>
  drop_na()

# calculate mean and maximum values and rename Lake
df_rdiv <- rdiv |> 
  group_by(fLake) |> 
  mutate(mean_rdiv = mean(rdiv)) |> 
  mutate(max_rdiv = max(rdiv)) |> 
  mutate(mean_sign = mean(sign)) |> 
  mutate(max_sign = max(sign)) |> 
  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
  rename(Lake = fLake)

# combine lake properties, species subgroups and response diversity

df_lm <- merge(df_rdiv, endemism_lake)

str(df_lm)

# phosphorus values as numeric
df_lm$Phos_max <- as.numeric(df_lm$Phos_max)

###Species richness and eutrophication
Multiple linear regression to calculate the effect of species richness and
eutrophication on response diversity.

###Fig. 3

```{r}
# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
# lm(mean_rdiv  ~ species richness + eutrophication)
# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
#   prepare these two plots

# Phosphorus data on Poschiavo is not available
df_lm_excl <- df_lm |> 
  filter(Lake != "Poschiavo")

# first lm for species richness ~ maximum phosphorus

lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm1)
shapiro.test(resid(lm1))
lmtest::bptest(lm1)
plot(lm1)

# model assumptions validated

# manually predict data
new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model1 <- cbind(new_data, prediction_lm1) |> 
  mutate(category ="color")

# second lm for maximum phosphorus ~ species richness
lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm2)
shapiro.test(resid(lm2))
lmtest::bptest(lm2)
plot(lm2)

# new data
new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model2 <- cbind(new_data2, prediction_lm2) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "#78BE63")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm1_plot <- bind_model1 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm2_plot <- bind_model2 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Maximum total phosphorus (µg/L)") +
  ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_3 <- ggarrange(lm1_plot, lm2_plot)

figure_3
# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

###Divergence figure 3

```{r}
# first lm for species richness ~ maximum phosphorus

lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm3)
shapiro.test(resid(lm3))
lmtest::bptest(lm3)
plot(lm3)

# model assumptions validated

# manually predict data
new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model3 <- cbind(new_data3, prediction_lm3) |> 
  mutate(category ="color")

# second lm for maximum phosphorus ~ species richness
lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm4)
shapiro.test(resid(lm4))
lmtest::bptest(lm4)
plot(lm4)

# new data
new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model4 <- cbind(new_data4, prediction_lm4) |> 
  mutate(category ="color")

# Plot Fig. 3
mycolor <- c("color" = "#88D44F")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm3_plot <- bind_model3 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm4_plot <- bind_model4 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Maximum total phosphorus (µg/L)") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_divergence <- ggarrange(lm3_plot, lm4_plot)

figure_divergence
# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_divergence)
# 
# # Closing the graphical device
# dev.off()

```
###Community subgroups
####Fig. 4a
We fitted simple linear models for the number of species in each category and the mean dissimilarity in each lake. 
```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)

# checking model assumptions
summary(lm_endemic)
shapiro.test(resid(lm_endemic))
lmtest::bptest(lm_endemic)
plot(lm_endemic)

new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))

prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")

df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)

# checking model assumptions
summary(lm_nn)
shapiro.test(resid(lm_nn))
lmtest::bptest(lm_nn)
plot(lm_nn)

new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")

df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

lm_nne <- lm(mean_rdiv ~ native, data = df_lm)

# checking model assumptions
summary(lm_nne)
shapiro.test(resid(lm_nne))
lmtest::bptest(lm_nne)
plot(lm_nne)

new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")

df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

# Translocated (non-native to region) species

lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)

# checking model assumptions
summary(lm_trans)
shapiro.test(resid(lm_trans))
lmtest::bptest(lm_trans)
plot(lm_trans)

new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")

df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)

mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")

plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"),
                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
   scale_linetype_manual(values = c("endemic" = "solid",
                                   "non_native" = "dashed",
                                   "non_endemic_native" = "solid",
                                   "translocated" = "dashed"), labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)"))

plot_all_lms
# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()
```

####Divergence Fig. 4a

```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)

# checking model assumptions
summary(deriv_lm1)
shapiro.test(resid(deriv_lm1))
lmtest::bptest(deriv_lm1)
plot(deriv_lm1)

new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))

prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")

df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)

# checking model assumptions
summary(deriv_lm2)
shapiro.test(resid(deriv_lm2))
lmtest::bptest(deriv_lm2)
plot(deriv_lm2)

new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")

df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)

# checking model assumptions
summary(deriv_lm3)
shapiro.test(resid(deriv_lm3))
lmtest::bptest(deriv_lm3)
plot(deriv_lm3)

new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")

df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

# Translocated (non-native to region) species

deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)

# checking model assumptions
summary(deriv_lm4)
shapiro.test(resid(deriv_lm4))
lmtest::bptest(deriv_lm4)
plot(deriv_lm4)

new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")

df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)

mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")

lm_divergence_all <- df_deriv_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1, linetype = "dashed") +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)

lm_divergence_all

# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(lm_divergence_all)
# 
# dev.off()
```
#### Fig. 4b 
Predictions of all species shown inside the categories. 
```{r, fig.show='hide'}

# Plot species predictions inside categories across all lakes
# 5 categories;
# endemic: geographically constrained range
# non-native: from NA or Asia 
# non-endemic native: native to Switzerland but not endemic
# non-native region: native to Switzerland and surroundings but has been translocated
# to other Swiss lakes where the species was not native
# endemic translocated: endemic species that were translocated to other lakes

# read in table with the categories and rename the categories
species_category <- read_excel("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
  mutate(category = ifelse(category == "native", "non_endemic_native", category)) |> 
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_category$category <- as.factor(species_category$category)
levels(species_category$category)

# merge predictions with categories
model_pred_categories <- merge(model_predictions, species_category)

# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# Rescale the predictions to 0-1 
model_pred_categories_rescaled <- model_pred_categories |> 
  group_by(species) |> 
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

# Plot the rescaled predictions
plot_pred <- model_pred_categories_rescaled |>
  # exclude Lepomis gibbosus
  filter(species != "Lepomis_gibbosus") |> 
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line(linewidth = 0.8) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

# save plot as TIFF
# tiff(paste("total_models/plots/plot_category_predictions.tiff", sep = ""), units="in", width=15, height=4, res=300)
# 
# plot(plot_category_predictions)
# 
# dev.off()

>>>>

---
title: "Models, derivatives, response diversity"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---
#2. Data Analysis
##General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@

# function for the temperature only GAMs
source(here("function_GAM.R"))

# function for temp and depth models 
source(here("function_temp_depth.R"))
```

@@

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions
# with three subfolders: temp, temp_test, depth_temp (total_models/predictions_depth_temp)

df_exclusion <- df_4 |> 
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", "Lepomis_gibbosus"))

# all temperature models, no simplificaiton

predictions(df_exclusion)

predictions(df_1)
predictions(df_2)
predictions(df_3)
predictions(df_4)

# temperature models with simplified models
predictions_temp(df_1)
predictions_temp(df_2)
predictions_temp(df_3)
predictions_temp(df_4)

# sensitivity
predictions_depth_temp(df_1)
predictions_depth_temp(df_2)
predictions_depth_temp(df_3)
predictions_depth_temp(df_4)

@@

df_predictions_all <- list.files(path = "total_models/depth_temp/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

@@

  # model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
  #                  s(Depth_sample, k = 3)
  #                           + s(fLake, bs = 're')
  #                           +  s(fProtocol, bs = 're'), family = ziP())
  # print(glance(model_0))

@@

AIC_testing(total_models, "Alburnus_arborella")
# 8 models are run with binomial temp and depth

@@
